Assessment of the KLGO Coastal Waterbird Survey Monitoring Program

### 2018-09-12 analysis/paper/paper.Rmd
Just put an R in Swarthmore :)...is it supposed to be able to knit? There's an error but not sure what it means.
tiny edits to start clarifying section headers, ID initial diagnostics for Madeleine to pursue

---
title: "Assessment of the KLGO Coastal Waterbird Survey Monitoring Program"
author:
  - name: Joel H. Reynolds^[U.S. National Park Service, Alaska Region, Natural Resources Science Team]
  - name: Madeleine Ward^[Swathmore College, Dept. of Mathematics and Statistics]
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: word_document

@@

Recommendations:

@@

Define the Target Audiences for the information and their Objectives in using the information
Decision Makers
Who are target audiences for this information?
Who uses it or may use it?
Local contact: C. Elaine Furbish, Skagway Bird Club, Google Site
Predominantly historical park; limited resource staff and capacity
Visible natural resource of interest to local community and visitors; interp and educational opportunity; topic of information requests; provides ‘community service’ given absence of any other agency with natural resource staff;
Annual report (time lag?); National Resource Condition Assessment; other? State of the Park?
Potential for contribution to regional, flyway, and WASO programs (e.g., CCRP & Audubon climate watch, etc.)
Current sightings passed to local bird club & used by avian-interested visitors
Used in quarterly regional ‘significant bird observations’, which in turn is used to develop the Alaska-wide summary for the journal North American Birds (published by the American Birding Association http://www.aba.org/nab ) and includes ‘...sightings of birds that are out of range or out of season or occur regularly in small numbers, noteworthy breeding records, unusually large or small numbers of a particular species, unusual migration dates, etc….’ (S. C. Heinl, 2017, Summary of SE AK Bird Observations, June-July 2017, accessed 2018 June 6).
Queries to Heather Renner & Robin Corcoran re: PSG-type users, etc.
Opportunity for community based monitoring to track projected spp impacts of climate change, change in phenology, occurrence, abundance, etc.?

Decisions
What types of decisions is it / will it be used for?
(specify for each target audience/ end user)
No specific decisions
Current:
KLGO - spp lists, phenology of occurrence (guidance to visitors), trends in richness (NRCA), abundance?
Skagway bird club - spp list, current occurrences (activity re: rare & unusual)
SE AK / Regional - spp lists; unusual sightings, etc. (spp occurrences & broad seasonal timing).
AKR -
WASO/NPS - ??CCRP?? Question to Gregor & Mel (AK Audubon). Could be used for garnering community engagement in Climate Watch & other national efforts associated with Audubon’s Climate Report. Not used in continental analyses.
“One key message that I want to emphasize is that the focus of this work was purely on climate suitability.  What this means, as we indicate in the briefs’ “Important” box up front and “Caveats” section further back, is that significant changes in climate suitability, as measured here, will not always result in a species response, and all projections should be interpreted as potential trends. Multiple other factors mediate responses to climate change, including habitat availability, ecological processes and biotic interactions, dispersal capacity, species' evolutionary adaptive capacity, and phenotypic plasticity (e.g., behavioral adjustments). Ultimately, models can tell us where to focus our concern and which species are most likely to be affected, but monitoring is the only way to validate these projections and should inform any on-the-ground conservation action.” Gregor Schuurman, email to DENA

@@

Phase 2 - Monitoring & Analysis Design
HOW monitor? Clarify the various design choices.

@@

Survey Design 
Species: 
Targets: loons, grebes, cormorants, herons, ducks, geese, plovers, sandpipers, gull, alcids, [kingfishers]
Questions: clarify role of kingfishers.

@@

Target Frame - space, time, scale, target populations
Spatial: Upper Lynn Canal visible from coastline from Skagway to Dyea, including Skagway & Taiya river mouths and Nahku Bay (between Skagway and Dyea Point)

@@

Temporal: ??how define ‘season’??
Questions: what is cue to start of survey season?
Diagnostic analysis => check initial counts x spp

Sample Frame - 
Spatial: same coverage as Target Frame. Area split into eight survey areas (see Hahr & Trapp 2004).
Questions: Access to all units? Ease of observation of different units? Ease of distinguishing boundaries?
Diagnostic analysis => value of all units - only some spp in unit X?; “unit by species occurrence” relationships?
Sample Selection process - any strata or clusters? Two-stage processes? Probabilistic selection? [get in to enumerative vs analytical distinction of Deming?]
Currently survey all of frame (‘census’ of units); no spatial sampling uncertainty. 
All spatial uncertainty comes from Observation Process and detection issues.
Temporal sampling uncertainty remains.

@@

Survey timing 
Per season -

@@

Observation Process
Question: 
How determine boundaries? Visual markers.
How handle movement of birds into and out of unit? Check SOPs but basically ‘until things stop’. Problem is that that is an accumulating approach that will bias high areas with lots of productive habitat and/or well timed relative to changing tidal conditions and sequencing of survey.
How handle tendency to record everything of note: e.g., recording observations in neighboring units when really supposed to be focused on this unit. E.g., variation in time observing and effort observing across units & sampling events. This fails to track actual observation activity (time) devoted to each unit as well as preferential bias to overestimate use in areas. Think about Forsell & Zweifelhofer?  Chat w/ Emily & Mark Otto & Andy?
Covariates - captured & not captured. Value, use in analyses; more reliable alternative data sources; alignment of spatial scales of interest and time scales of interest.
Lack of control over time at each site; portion spent traveling between observation points, variation in number of observation points for each site.
Variation in survey effort across sites, across per unit area, across surveys (days), etc.
Major weather conditions (Northerlies, Southerlies) - simplify conditions to these
Unit 1 - recording and impact of presence & # of cruise ships in dock, helicopter tours, etc., etc., other human disturbances.
Distinguish value of covariates for act of (i) presence/absence assessment, (ii) abundance count, (iii) composition & breeding state.
Check SOP to clarify categorization of conditions.
Temp and wind greatly burden data collection process (slowing down greatly).
Variation in observer #, training, and skill across surveys, across years

Diagnostic analyses =>
.distribution of time duration at each site by date and year
.correlation across units?
.average surface area per unit
.something about sequencing of observation sites (e.g., confounding with tidal state)?
.distribution of observers by survey date by year
.distribution of observer experience(?)  by survey date by year?

Detection issues associated with survey conditions
.species specific counts w/in season by weather conditions and wave conditions
.species occurrence by wave conditions & weather conditions
.species age and composition by wave conditions & weather conditions 
.total # species by weather conditions, wave conditions; internally consistent contrasts to demonstrate value, or lack thereof, of current covariates.
.richness by major weather conditions (Northerlies, Southerlies, etc.) and for each spp
.distribution of weather conditions
.distribution of survey time by weather conditions
.diagnostic to ID key survey times to ensure high quality observers?
.consistency in ‘breeding periods’ across years (as captured in data)
.quick diagnostics to ID covariates seldom recorded, not consistently recorded, consistently valued (e.g., no basis for contrast), etc.

@@

Species identification issues
Priority ID concerns:
Commonly encountered species that can be difficult
Work with Elaine & Heather & Laura to develop list of commonly confused species based on those commonly encountered (database or Skagway Bird Club sightings)
E.g., scoters, scaup, murrelets, etc.)
Target Spp of interest
Training materials or other improvements
Db error checking?
Other?
Other expected sources of bias
Species ID
Behavior classification
Changes in equipment
Summary Metrics
Confounding factors & Covariates

Analysis Design
Priority 
Summaries of interest 
(tie to Conceptual Model and Information Objectives)
Associated analyses
Reporting schedule
Questions: Who does the analyses? Interest in setting up as Reproducible Research draft report? 
General delay in reporting? When normally completed?

@@

Data Management & Reporting Structures / Software / Workflows
Data Models
Design Requirements / User reporting needs

>>>>

---
title: "Assessment of the KLGO Coastal Waterbird Survey Monitoring Program"
author:
  - name: Joel H. Reynolds^[U.S. National Park Service, Alaska Region, Natural Resources Science Team]
  - name: Madeleine Ward^[Swarthmore College, Dept. of Mathematics and Statistics]
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: word_document

@@

### Recommendations:

@@

Define the _Target Audiences_ for the information and their Objectives in using the information

_Decision Makers_
Who are target audiences for this information?
Who uses it or may use it?
Local contact: C. Elaine Furbish, Skagway Bird Club, Google Site
Predominantly historical park; limited resource staff and capacity
Visible natural resource of interest to local community and visitors; interp and educational opportunity; topic of information requests; provides ‘community service’ given absence of any other agency with natural resource staff;
Annual report (time lag?); National Resource Condition Assessment; other? State of the Park?
Potential for contribution to regional, flyway, and WASO programs (e.g., CCRP & Audubon climate watch, etc.)
Current sightings passed to local bird club & used by avian-interested visitors
Used in quarterly regional ‘significant bird observations’, which in turn is used to develop the Alaska-wide summary for the journal North American Birds (published by the American Birding Association http://www.aba.org/nab ) and includes ‘...sightings of birds that are out of range or out of season or occur regularly in small numbers, noteworthy breeding records, unusually large or small numbers of a particular species, unusual migration dates, etc….’ (S. C. Heinl, 2017, Summary of SE AK Bird Observations, June-July 2017, accessed 2018 June 6).
Queries to Heather Renner & Robin Corcoran re: PSG-type users, etc.
Opportunity for community based monitoring to track projected spp impacts of climate change, change in phenology, occurrence, abundance, etc.?

_Decisions_
What types of decisions is it / will it be used for?
(specify for each target audience/ end user)
No specific decisions
Current:
KLGO - spp lists, phenology of occurrence (guidance to visitors), trends in richness (NRCA), abundance?
Skagway bird club - spp list, current occurrences (activity re: rare & unusual)
SE AK / Regional - spp lists; unusual sightings, etc. (spp occurrences & broad seasonal timing).
AKR -
WASO/NPS - ??CCRP?? Question to Gregor & Mel (AK Audubon). Could be used for garnering community engagement in Climate Watch & other national efforts associated with Audubon’s Climate Report. Not used in continental analyses.
“One key message that I want to emphasize is that the focus of this work was purely on climate suitability.  What this means, as we indicate in the briefs’ “Important” box up front and “Caveats” section further back, is that significant changes in climate suitability, as measured here, will not always result in a species response, and all projections should be interpreted as potential trends. Multiple other factors mediate responses to climate change, including habitat availability, ecological processes and biotic interactions, dispersal capacity, species' evolutionary adaptive capacity, and phenotypic plasticity (e.g., behavioral adjustments). Ultimately, models can tell us where to focus our concern and which species are most likely to be affected, but monitoring is the only way to validate these projections and should inform any on-the-ground conservation action.” Gregor Schuurman, email to DENA

@@

## Phase 2 - Monitoring & Analysis Design
HOW monitor? Clarify the various design choices.

@@

### Survey Design 
Species: 
Targets: loons, grebes, cormorants, herons, ducks, geese, plovers, sandpipers, gull, alcids, [kingfishers]
Questions: clarify role of kingfishers.

@@

*Target Frame* - space, time, scale, target populations
*Spatial* Upper Lynn Canal visible from coastline from Skagway to Dyea, including Skagway & Taiya river mouths and Nahku Bay (between Skagway and Dyea Point)

@@

*Temporal* ??how define ‘season’??

Questions: what is cue to start of survey season?
*Madeleine later* Diagnostic analysis => check dates & unit of first non-zero counts each year for each spp (relative to dates of initiating surveys)

*Sample Frame* 
Spatial: same coverage as Target Frame. Area split into eight survey areas (see Hahr & Trapp 2004).
Questions: Access to all units? Ease of observation of different units? Ease of distinguishing boundaries?
*Madeleine later* Diagnostic analysis => value of all units - only some spp in unit X?; “unit by species occurrence” relationships?

*Sample Selection process* - any strata or clusters? Two-stage processes? Probabilistic selection? [get in to enumerative vs analytical distinction of Deming?]
Currently survey all of frame (‘census’ of units); no spatial sampling uncertainty. 
All spatial uncertainty comes from Observation Process and detection issues.
Temporal sampling uncertainty remains.

@@

*Survey timing* 
Per season -

@@

###Observation Process
Question: 
How determine boundaries? Visual markers.
How handle movement of birds into and out of unit? Check SOPs but basically ‘until things stop’. Problem is that that is an accumulating approach that will bias high areas with lots of productive habitat and/or well timed relative to changing tidal conditions and sequencing of survey.
How handle tendency to record everything of note: e.g., recording observations in neighboring units when really supposed to be focused on this unit. E.g., variation in time observing and effort observing across units & sampling events. This fails to track actual observation activity (time) devoted to each unit as well as preferential bias to overestimate use in areas. Think about Forsell & Zweifelhofer?  Chat w/ Emily & Mark Otto & Andy?
#### Covariates 
- captured & not captured. Value, use in analyses; more reliable alternative data sources; alignment of spatial scales of interest and time scales of interest.

Lack of control over time at each site; portion spent traveling between observation points, variation in number of observation points for each site.

Variation in survey effort across sites, across per unit area, across surveys (days), etc.

Major weather conditions (Northerlies, Southerlies) - simplify conditions to these
Unit 1 - recording and impact of presence & # of cruise ships in dock, helicopter tours, etc., etc., other human disturbances.

Distinguish value of covariates for act of (i) presence/absence assessment, (ii) abundance count, (iii) composition & breeding state.

Check SOP to clarify categorization of conditions.

Temp and wind greatly burden data collection process (slowing down greatly).
Variation in observer #, training, and skill across surveys, across years

*Madeleine early* Diagnostic analyses =>
.distribution of time duration at each site by date and year
.correlation across units?
.average surface area per unit
.something about sequencing of observation sites (e.g., confounding with tidal state)?
.distribution of observers by survey date by year
.distribution of observer experience(?)  by survey date by year?

#### Detection issues associated with survey conditions
.species specific counts w/in season by weather conditions and wave conditions
.species occurrence by wave conditions & weather conditions
.species age and composition by wave conditions & weather conditions 
.total # species by weather conditions, wave conditions; internally consistent contrasts to demonstrate value, or lack thereof, of current covariates.
.richness by major weather conditions (Northerlies, Southerlies, etc.) and for each spp
.distribution of weather conditions
.distribution of survey time by weather conditions
.diagnostic to ID key survey times to ensure high quality observers?
.consistency in ‘breeding periods’ across years (as captured in data)
.quick diagnostics to ID covariates seldom recorded, not consistently recorded, consistently valued (e.g., no basis for contrast), etc.

@@

####Species identification issues
Priority ID concerns:
Commonly encountered species that can be difficult
Work with Elaine & Heather & Laura to develop list of commonly confused species based on those commonly encountered (database or Skagway Bird Club sightings)
E.g., scoters, scaup, murrelets, etc.)
Target Spp of interest
Training materials or other improvements
Db error checking?
Other?
Other expected sources of bias
Species ID
Behavior classification
Changes in equipment
Summary Metrics
Confounding factors & Covariates

###Analysis Design
Priority 
Summaries of interest 
(tie to Conceptual Model and Information Objectives)
Associated analyses
Reporting schedule
Questions: Who does the analyses? Interest in setting up as Reproducible Research draft report? 
General delay in reporting? When normally completed?

@@

### Data Management & Reporting Structures / Software / Workflows
Data Models
Design Requirements / User reporting needs
